Type Refinements for the Working Class

### 2015-11-22 paper.tex
Entirely remove the ambient logical framework, higher order judgments
I think that things will be clearer if we are fully explicit about the
quantification over renamings, even if we could factor it out into the
presheaf machinery.

@@

\section{The Ambient Judgmental Framework}
\label{section:ambient-lf}

\ifcolored%
In this paper, we hint at the \emph{modes} of judgments and
assertions~\cite{harper:2016} using colors, marking inputs with
$\IMode{\texttt{blue}}$ and outputs with $\OMode{\texttt{red}}$. As a rule of
thumb, inputs are things which are supplied when checking the correctness of a
judgment, and outputs are things which are synthesized in the process. \fi

For any abt signature $\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$, we can
deploy a generic judgmental framework equipped with higher-order judgments. A
judgment is always defined using a \emph{meaning explanation}, which is a
specification of the conditions under which it may be asserted (sc.\ its
evidence) \cite{martin-lof:1996}.

\paragraph{Renaming convention} Because the evidence of a judgment will often
contain symbols $u,v,\dots$ we informally adopt the convention that the
evidence of a judgment shall always be subject to fresh renamings of its free
symbols (i.e.\ the validity of a judgment shall not depend on choice of names).
This apparatus is developed formally in Appendix~\ref{appendix:lf}, where the
judgmental framework is cast as a topos of covariant presheaves on symbol contexts
$\SCtx$.

\begin{definition}[Hypothetical Judgment]
  For two judgments $\mathcal{J}_1$ and $\mathcal{J}_2$, we can form the
  hypothetical judgment $\HypJ{\mathcal{J}_2}{\mathcal{J}_1}$, whose evidence
  shall be an effective transformation of any evidence for $\mathcal{J}_1$ into
  evidence for $\mathcal{J}_2$. It is important to keep in mind that this is a
  \emph{semantic} consequence and therefore expresses admissibilities, in
  contrast to the \emph{logical} consequence which expresses
  derivabilities~\cite{harper:2016}. We will not have any need for the latter.

  \paragraph{Notation}
  For a hypothetical judgment with multiple hypotheses, we will write
  $\HypJ{\mathcal{J}_3}{\mathcal{J}_1, \mathcal{J}_2}$ rather than
  $\HypJ{\mathcal{J}_3}{\mathcal{J}_1\times\mathcal{J}_2}$ or
  $\HypJ{(\HypJ{\mathcal{J}_3}{\mathcal{J}_2})}{\mathcal{J}_1}$.
\end{definition}

\begin{definition}[General Judgment]
  For a metavariable context $\IsMetaCtx{\Theta}$, we can define the set
  of its substitutions as follows:
  \[
    \Define{{\square\Theta}}{
      \prod_{(\MV{m}:v)\in\Theta}
        \MkSet{E\mid\IsAbtUnmoded{\cdot}{\cdot}{\cdot}{E}{v}}
    }
  \]

  For any $\square{\Theta}$-indexed family of judgments $\mathcal{J}$, we can
  express the general judgment $\GenJ{\Theta}{\mathcal{J}}$ as being evident
  when $\mathcal{J}(\vec{E})$ is evident for each
  $\Of{\vec{E}}{\square\Theta}$; in other words, the evidence of the general
  judgment is an effective transformation of $\Theta$-substitutions to their
  fibres in $\mathcal{J}$.  Recall that metavariables have \emph{valences}
  rather than sorts; as such, the general judgment allows us to quantify over
  terms of higher type.

   \paragraph{Notation}

   We will write $\HPGenJ{\Upsilon}{\Gamma}{\Theta}{\mathcal{J}}$ for
   $\GenJ{\Theta'}{\mathcal{J}}$, where
   $\Lookup{\Theta'}{\MV{m}}{\MkValence{\Upsilon,\vec{\sigma}}{\Gamma,\vec{\tau}}{\tau}}$
   if $\Lookup{\Theta}{\MV{m}}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}}$.
   Additionally, for clarity we will often write
   $\GenJ{\MV{m}:\sigma}{\mathcal{J}}$ rather than
   $\GenJ{\MV{m}:\MkValence{\cdot}{\cdot}{\sigma}}{\mathcal{J}}$; likewise, in
   the body of $\mathcal{J}$, when there are contextually salient
   $\vec{u}$, $\vec{x}$, we will simply write $\MV{m}$ rather than the more proper
   $\MApp{\MV{m}}{\vec{u}}{\vec{x}}$.

After this section, we will define a judgment either by rules, or by an
informal semantical explanation, in both cases leaving its intensional
character implicit. A judgment is said to be \emph{correct} or \emph{valid}
just in case it is globally inhabited.

@@

\newpage
\begin{appendices}
\section{Formal Definition of the Judgmental Framework}
\label{appendix:lf}

As noted earlier, we must be cautious in the definitions of both atomic and
higher-order judgments that their evidence shall always respect the
\emph{renaming convention}. We do so explicitly in this appendix by developing
our judgmental framework as a presheaf topos over symbol contexts.

A judgment is, then, an object in the presheaf category
$\Define{\mathbb{J}}{\Sets^\SCtx}$; in other words, judgments are identified
with the intensional set of its renameable evidence. $\mathbb{J}$, like all
presheaf categories, is a topos. The task at hand, then, is to show how to
replace the informal meaning explanations that we gave in
section~\ref{section:ambient-lf} with precise definitions in the topos
$\mathbb{J}$.

\subsection{Hypothetical Judgment}

Hypothetical judgment $\HypJ{\mathcal{J}_2}{\mathcal{J}_1}$ is defined using the
exponential construction~\cite[p.~46]{maclane-moerdijk:1992}, which expresses
the semantic consequence of $\mathcal{J}_2$ from $\mathcal{J}_1$:
\begin{align*}
  \ADefine{
    \left(\HypJ{\mathcal{J}_2}{\mathcal{J}_1}\right)(\Upsilon)
  }{
    \left({\mathcal{J}_2}^{\mathcal{J}_1}\right)(\Upsilon)
  }\\
  &\cong
  \OMode{
    \Hom{\mathbb{J}}{
      \Yoneda{\Upsilon}\times\mathcal{J}_1
    }{
      \mathcal{J}_2
    }
  }\\
  &\cong
  \OMode{
    \int_{\Upsilon'}{{\Upsilon\inj\Upsilon'}
      \Rightarrow{{{\mathcal{J}}_{2}(\Upsilon')}^{{\mathcal{J}}_{1}(\Upsilon')}}}
  }
\end{align*}

In other words, the hypothetical judgment $\HypJ{\mathcal{J}_2}{\mathcal{J}_1}$
is evident when there shall ``forevermore'' (i.e. for any applicable renaming
of symbols) be an effective transformation of evidence for the antecedent into
evidence for the consequent.

\subsection{General Judgment}

For any diagram $\Of{\mathcal{J}}{\mathbb{J}^{\square\Theta}}$, we can define
the general judgment $\GenJ{\Theta}{\mathcal{J}}$ as follows:
\begin{align*}
  \ADefine{
    \left(\GenJ{\Theta}{\mathcal{J}}\right)(\Upsilon)
  }{
    \Hom{\mathbb{J}}{
      \Yoneda{\Upsilon}
    }{
      \varprojlim\mathcal{J}
    }
  }\\
  &\cong
  \OMode{
    \int_{(\Upsilon',\,\vec{E})}{\Upsilon\inj\Upsilon'\Rightarrow{\mathcal{J}(\vec{E})(\Upsilon')}}
  }
\end{align*}

 Intuitively, the general judgment $\GenJ{\Theta}{\mathcal{J}}$ is evident
 when the family of judgments $\mathcal{J}$ shall forevermore have a global
 section (i.e.\ for any substitution $\Member{\vec{E}}{\square\Theta}$,
 $\mathcal{J}(\vec{E})$ shall be evident).

\subsection{Internal Judgments}
It often happens that the evidence for a judgment is the knowledge of the
meaning of a new form of judgment; this is the case when we say ``To know
$\mathcal{J}$ is to know \emph{the meaning of} another judgment
$\mathcal{J}'$.'' This can be neatly expressed in our topos using the subobject
classifier $\Omega$, where the evidence of $\Omega$ at $\Upsilon$ gives rise to
an \emph{internal judgment} which is defined at all $\Upsilon'$ for which we
have $\Of{\rho}{\Upsilon\inj\Upsilon'}$.

A \emph{global internal (family of) judgments} is a section
$\Of{J}{\int_\SCtx {\Omega^{\mathcal{J}}}}\cong
\IMode{\int_\Upsilon\Hom{\mathbb{J}}{\Yoneda{\Upsilon}\times\mathcal{J}}{\Omega}}$,
which can be externalized as a ``total'' judgment $\Of{\Embed{J}}{\mathbb{J}}$
for whose derivations at $\Upsilon$ are the arrows
$\Of{\phi}{\Hom{\mathbb{J}}{\Yoneda{\Upsilon}}{\mathbb{1}\times\mathcal{J}}}$
such that the following diagram commutes:
\[
  \begin{tikzcd}[cramped,sep=huge]
    \IMode{\Yoneda{\Upsilon}}
      \arrow[r, dashed, "\OMode{\phi}"]
      \arrow[rd, "\IMode{J_\Upsilon\Pair{1}{\pi_2\circ\phi}}"']
    & \IMode{\mathbb{1}\times\mathcal{J}} \arrow[d, "\IMode{\mathbf{true}\circ\pi_1}"] \\
    & \IMode{\Omega}
  \end{tikzcd}
\]

Then, to say $\IMode{\Upsilon}\parallel\Embed{J}(E)$ where
$\Member{E}{\mathcal{J}(\Upsilon)}$ is to assert that we have evidence
$\Member{F}{\Embed{J}(\Upsilon)}$ such that $\IsEq{\pi_2(F)}{E}$.

\end{appendices}

>>>>

\ifcolored%
\paragraph{Colors}
In this paper, we hint at the \emph{modes} of judgments and
assertions~\cite{harper:2016} using colors, marking inputs with
$\IMode{\texttt{blue}}$ and outputs with $\OMode{\texttt{red}}$. As a rule of
thumb, inputs are things which are supplied when checking the correctness of a
judgment, and outputs are things which are synthesized in the process. \fi

@@

@@
